Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews

### 2025-07-07 analysis/paper/paper.qmd
Added 1 chunck of code, changed Shannon's index, modified text for h-indices
1) I added the chunk of code found in the version 1.1 of the quarto document since it is the only place where there is the code to reproduce entirely the firue 3 of the published paper (figure plot over time). It does not recreate exactly the same plot but it is very close. While in the v1.3 it calls directly the .png but we can not recreate the figure.

2) I modified the calculation of the Shannon's index so that it is the same as in Fanelli & Glanzel 2013. It should indeed be the diversity of *sources* and not of *r

While Fanelli and Glänzel (2013) analysed papers published in a single year (2012), I found only `r n_items_df_2012` papers for that same year, and `r round(prop_pub_after2012*100, 0)`% of papers in the sample published after that date. To make efficient use of the available data and ensure robust representation from different areas of archaeology, including those with lower frequencies of journal article publication, I analysed 9697 papers published during 1975-2025. This sample was collected from Clarivate’s Web of Science database by first selecting the Web of Science category ‘Archaeology’ and the Document type ‘article’ (n = 28,871). To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate’s Journal Citation Indicator. Finally, I excluded journals with less than 100 articles in the database, resulting in `r n_journals` journals.

@@

# compute diversity of all citations for each article (habitat)
shannon_per_item <- 
  ref_list7 %>% 
  group_by(id, x) %>% 
  tally() %>% 
  mutate(n_in_article = n) %>% 
  select(-n) %>% 
  left_join(all_cited_items) %>% 
  mutate(p_i = n / sum(n, na.rm = TRUE)) %>% 
  mutate(p_i_ln = log(p_i)) %>% 
  group_by(id) %>% 
  summarise(shannon = -sum(p_i * p_i_ln, na.rm = TRUE)) %>% 
  mutate(id = as.numeric(id)) %>% 
  arrange(id)  %>% 
  left_join(items_df)

@@

@@

>>>>

While Fanelli and Glänzel (2013) analysed papers published in a single year (2012), I found only `r n_items_df_2012` papers for that same year, and `r round(prop_pub_after2012*100, 0)`% of papers in the sample published after that date. To make efficient use of the available data and ensure robust representation from different areas of archaeology, including those with lower frequencies of journal article publication, I analysed 9697 papers published during 1975-2025. This sample was collected from Clarivate’s Web of Science database by first selecting the Web of Science category ‘Archaeology’ and the Document type ‘article’ (n = 28,871). To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their 2022 Impact Factor as reported by Clarivate’s Journal Citation Indicator. Finally, I excluded journals with less than 100 articles in the database, resulting in `r n_journals` journals.

@@

# compute diversity of all citations for each article (habitat)
shannon_per_item <- 
  ref_list7 %>% 
  group_by(id, journal_name) %>% 
  tally() %>% 
  group_by(id) %>% 
  mutate(p_i = n / sum(n, na.rm = TRUE)) %>% 
  mutate(p_i_ln = log(p_i)) %>% 
  group_by(id) %>% 
  summarise(shannon = -sum(p_i * p_i_ln, na.rm = TRUE)) %>% 
  mutate(id = as.numeric(id)) %>% 
  arrange(id)  %>% 
  left_join(items_df)

@@

library(ggplot2)
library(tools)
library(stringr)
library(purrr)

@@

```{r}
#| label: fig-change-over-time_from_V1_1
#| fig-cap: Distribution of article characteristics for archaeology articles over time. Data points represent individual articles. The colour of the points indicates if the overall trend is toward softer (orange) or harder (green) Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series. 
#| fig-height: 4

library(ggpmisc)
library(mgcv)
formula <-  y ~ x

over_time_long_colour_gams <- 
over_time_long_colour %>% 
  nest(.by = variable) %>% 
  mutate(mod_gam = lapply(data, 
                          function(df) gam(year ~ s(value, bs = "cr"), 
                                           data = df)))

over_time_long_colour_gams_summary <- 
over_time_long_colour %>% 
  nest(.by = variable) %>% 
  mutate(fit = map(data, ~mgcv::gam(year ~ s(value, bs = "cs"), data = .)),
         results = map(fit, glance),
         R.square = map_dbl(fit, ~ summary(.)$r.sq)) %>%
  unnest(results) %>%
  select(-data, -fit) %>% 
  select(variable, adj.r.squared)

over_time_long_colour_gams_summary_df <- 
over_time_long_colour %>% 
  left_join(over_time_long_colour_gams_summary)

plot = ggplot() + 
  geom_point(data = over_time_long_colour_gams_summary_df, 
       aes(year,  
           value, 
           colour = becoming_more_scientific),
       alpha = 0.5) + 
  geom_smooth(data = over_time_long_colour_gams_summary_df,
              aes(year, value),
              method="gam", 
              formula = y ~ s(x, bs = "cs"),
              se = FALSE,  
              size = 2, 
              colour = "#7570b3") +
  facet_wrap( ~ variable,
              scales = "free_y") + 
  theme_bw(base_size = base_size) +
  scale_color_manual(values = c("#d95f02",
                                "#1b9e77" )) +
  guides(colour = "none") +
  ylab("") +
  geom_text(data = over_time_long_colour_gams_summary_df %>% 
              group_by(variable) %>% 
              summarise(max_value = max(value),
                        adj.r.squared = unique(adj.r.squared)),
           aes(
           x = 1980, 
           y = max_value, 
           label = paste("Pseudo R² = ", 
                         signif(adj.r.squared, 
                                digits = 3))),
           hjust = 0, 
           vjust = 1.5,
           size = 2)
